Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
more comments

% WL: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis. However, the reader will still be mostly focused on phenotypic heterosis, and more may need to be done to refocus on the transcriptome. Do we have good real-world examples of heterosis in the expression important or interesting genes? This may be a good question for Dan.

@@

@@

@@

@@

@@

@@

@@

@@

As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.

>>>>

% WML: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis. However, the reader will still be mostly focused on phenotypic heterosis, and more may need to be done to refocus on the transcriptome. Do we have good real-world examples of heterosis in the expression important or interesting genes? This may be a good question for Dan.

@@

% WML: Maybe it would be good to quickly clarify what a formal heterosis RNA-seq experiment is.

@@

% WML: "borrowing information" is intuitive and informal. I agree that the term should be used, but there might be a slightly more appropriate use. Not a big deal.

@@

% WML: We were debating whether to use g for gene or f for feature, and either choice would work. But we use g, we should probably also use G to denote the number of genes, rather than F.

@@

% WML: Are the sequencing depths c_{vi} or \delta_{vi}? I think c_{vi} is better because \delta is used in the phi-alpha-delta parameterization of gene expression effects.

@@

% WML: Is \mu_g the vector of \mu_{gv} values? If so, please clarify.

@@

% WML: I assume the "hierarchical means" are the hyperparameters in the \eta vector and the "marginally uniform correlations" are the ones derived from the off-diagonal covariance elements of \Sigma. Small clarifications might be helpful in passing.

@@

% WML: The use of \phi in this section conflicts with the phi-alpha-delta parameterization in other heterosis models. Maybe you could replace phi, alpha, and delta by other letters in the ShrinkBayes section to avoid this conflict.

@@

% WML: \ref{s:intro} or \ref{s:heterosis}?

% WML: the first two sentences are unclear. I would delete the first sentence altogether and compare "varieties" to treatment groups or blocks for the sake of clarity.

As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$,

% WML: For mid-parent heterosis, you need \mu_{g3} \ne \frac{\mu_{g1} + \mu_{g2}}{2}. I don't know whether you need \mu_{g1}\le \mu_{g3}\le \mu_{g2} or \mu_{g2}\le \mu_{g3}\le \mu_{g1}, and I have doubts. It might be good to check with Dan on this.

mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.

% WML: You may want to clarify that the genes in the list are ranked by the strength of EH and then justify the lists's utility for geneticists and in the comparison study to follow. This could be a good segue.
